\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-04-30 Latex/main.tex
reorganization of the paper and new results in simulation study

Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in equation \ref{eq: RACVM equation}. This effect is estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship. This choice of covariate is motivated by \cite{heide-jorgensen_behavioral_2021}. A response model is compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model.

@@

The diffusion models are discussed in Section 2
Then,  Section 3 is dedicated to a comprehensive simulation study, whereas in the last section,we give an overview of the narwhal data available for the analysis and the results of the application of our models to the narwhal data.

@@

The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation. First, we recall the standard diffusion model in this context, and then we modify it to include the effect of the shore on the movement.

@@

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. Then the Euler-scheme is used to get the transitions of the location process. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, both the velocity and the position processes have a closed form formula. Though this model is discussed in \cite{gurarie_correlated_2017}, we found no mention of these explicit formulas. \\

@@

By integrating the velocity equation over time one gets the following conditional distribution for the location process :
\begin{equation}
	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
\end{equation}
Here, the covariance matrix $Q(\Delta)$ has an explicit form.

\[
Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
\]
The location process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
Both components of the location and velocity processes are independent.
and one can check that for $\omega=0$, the formulas match the ones of the  CTCRW model \cite{johnson_continuoustime_2008}. Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm and phase. Derivation of the formulas given in this paragraph is detailed in appendix~\ref{section: detailed RACVM computations}.\\

@@

\subsubsection{State-space model}
In most applications, it is needed to consider measurement errors through a state-space model. We suppose that positions are observed at discrete times $t_j$, $j \in \{1,\cdots, n\}$. $Y_j$ is the random vector representing the observed position at time $t_j$, while $X_j=X(t_j)$ is the true position at time $t_j$.

@@

Here, we consider parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time.
We make the approximation that these parameters are constant on each time step $[t_j,t_{j+1}]$ . Denote $\Delta_j=t_{j+1}-t_j$, $V_j=V(t_j)$, $V_j=V(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix} 
	\frac{1}{\tau_j} & -\omega_j \\
	\omega_j & \frac{1}{\tau_j}
\end{pmatrix}$.

\begin{equation}
	\forall j \in \{1, \cdots, n\} \quad 
	\left\{ \begin{array}{l}
		V_{j+1}=\mu_j +\exp(-A_j\Delta_j)(V_j-\mu_j)+\zeta_j \\
		X_{j+1}=X_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(_Vj-\mu_j)+\xi_j
	\end{array}\right. \\
	\label{eq: RCVM autoregressive formula}
\end{equation}

@@

One can check that for $\omega_j=0$, the formulas match the ones of the CTCRW model \cite{johnson_continuoustime_2008}.
This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter. This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} the special case $\omega=0$. Therefore, we extend this inference method to the more general model~\ref{eq: RACVM equation}.

@@

Here, $A(X(t),V(t))=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega(t) \\
	\omega(t) & \frac{1}{\tau}
\end{pmatrix}$
where $\omega(t)=f_{\omega}(D_{shore}(t),\Theta(t))$, $D_{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D_{shore}(s) \simeq D_{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
In this case, the formulas of section~\ref{section: RACVM} are still valid.
The question still hanging is : what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$ : this is the diameter of rotation. For the process to be constrained, one condition would be that $\rho$ always remains significantly lower than $D_{shore}$. We hint that the condition should also include the regularity of the domain, but this will be a subject for further work. Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D_{shore}$ approached $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate. 
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
A weight based on the distance to the shore can be defined as follows :
\[c(D_{shore})=\exp\left(-\kappa \times \left(\frac{D_{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D_{shore}$ can be formulated as follows :

\begin{equation}
f_{\omega}(\theta,D_{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D_{shore})
\label{eq: smooth omega}
\end{equation}
for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. The general model is very flexible, and different species-specific behaviours could be described. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. \\
Figure~\ref{fig: trueomega} shows a smooth function obtained from~\ref{eq: smooth omega} for the parameter $\omega$. Figure~\ref{fig: splineomega} shows three spline basis functions and a smooth function $f_{\omega}$ obtained as a combination of these basis functions.

\begin{figure}[H]
	\centering
		\centering
		\includegraphics[scale=0.25]{images/simulation study/true_smooth_omega_rect.png}
		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
			(a) $D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
		\label{fig: trueomega}
\end{figure}

\begin{figure}[H]
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.12]{images/simulation study/spline_smooth_omega_rect.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.12]{images/simulation study/spline_smooth_omega_rect.png}
		\caption{}
	\end{subfigure}
	\caption{}
	\label{fig: splineomega}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/illustrative_sample_rect.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/illustrative_sample_fjords.png}
		\caption{}
	\end{subfigure}
	\caption{Samples of the CRCVM in different domains. x and y axis are in metres and initial positions are indicated by a red cross and red point represent the nearest points on the boundary of the domain (a) One sample three day sample with time step $h=1$ min in a rectangular domain (b) One sample three day sample with time step $h=1$ min in Scoresby Sound fjords system in South-East Greenland}
	\label{fig: samples_rect_CRCVM}
\end{figure}

\section{Simulation study}

@@

\subsection{Simulation of constrained motion}
In order to test the constrained models we simulated the CRCVM process within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D_{shore}$.\\

$3$ days long trajectories were sampled in a rectangular domain with sides of length $8$ km from the CRCVM with the smooth function for $\omega$ displayed in figure~\ref{fig: trueomega}(a) and other parameters kept constant ($\nu = 4$ km/h, $\tau=1$ h). For each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary. Time steps were chosen equal to $1$ min for high frequency and subsampled to $5$ minutes for low frequency. Gaussian measurement errors with low standard deviation $\sigma_{obs}=3$ m and large standard deviations $\sigma_{obs}=30$ m were added to provide the actual observations. Of these $100$ trajectories, none reach the boundary for $\sigma_{obs}=3$ m and only one for $\sigma_{obs}=30$m. However, too small values of $D_0$ and $\omega_0$, as well as too large measurement errors may produce samples that reach the boundary of the domain. Figure~\ref{fig: samples_rect_CRCVM} shows some of the 100 samples obtained in the rectangular domain. \\

10 trajectories were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints. The parameter $\omega$ was defined as the smooth function displayed in~\ref{fig: trueomega}(b). Again, for each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore. Time steps were also kept constant, equal to $1$ min for high frequency and subsampled to $5$ min for low frequency. None of the 10 trajectories reached land. Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often. For the fjords system, this already happens if we try to sample directly with $5$ min time steps. Constrained samples within the geometry of Scoresby sound fjords are shown in~\ref{fig: samples_fjords_CRCVM}. Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.

\subsection{Estimation of angular velocity $\omega$ with tensor splines}

The samples from previous section were used to estimate the function $\omega$, both in the rectangular domain and in Scoresby Sound fjords system. Covariates $E_{shore}$ and $\Theta$ were computed from the observed positions and the smooth parameter $\omega$ was estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}. The marginal spline degree of freedom was set to $5$ for each covariate, which corresponds to $3$ knots in the splines. Hence, $25$ coefficients (including the intercept) were to be estimated. The other parameters of the diffusion were supposed to be known, as well as the measurement error. Underestimation of the smooth parameter $\omega$ is observed close to the shore as $\Theta$ approaches $\pm \pi$ (movement toward the shore). This is easily seen in figures \ref{fig: marginal_estimates_rect_CRCVM}(a)-(b) and and \ref{fig: marginal_estimates_fjords_CRCVM}(a)-(b).

As measurement error grows, the covariates that are estimated from the observed positions are becoming less accurate, which makes estimation more challenging. THta is why slight differences in figures ~\ref{fig: marginal_estimates_rect_CRCVM}(a)-(b).

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/omega_crcvm_rect_hf_ne1.png}
		\caption{}
	\end{subfigure}
	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in the rectangular domain. (a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
	\label{fig: estimate_rect_CRCVM}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_close.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_close.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_ne1_omega_far.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/crcvm_rect_hf_e1_omega_far.png}
		\caption{}
	\end{subfigure}
	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only, in the rectangular domain. The red curve represents the true value of $\omega$. (a)-(b): Fixed distance $D_{shore}=0.5$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m. (c)-(d): Fixed distance $D_{shore}=3$ km with respective measurement errors $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
\label{fig: marginal_estimates_rect_CRCVM}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{}
		\caption{}
	\end{subfigure}
	\caption{Estimates of the smooth parameter $\omega$ as a function of $D_{shore}$ and $\Theta$ in Scoresby Sound fjords system. (a)-(b): Estimation from data with time step $h=1$ min (high frequency) with respective measurement errors standard deviation $\sigma_{obs}=3$ m and $\sigma_{obs}=30$ m.}
	\label{fig: estimate_fjords_CRCVM}
\end{figure}

>>>>

Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in equation \ref{eq: RACVM equation}. This effect is estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship. This choice of covariate is motivated by \cite{heide-jorgensen_behavioral_2021}. A response model is compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model parameters.

@@

The diffusion models are discussed in Section 2 and simulation study results are shown.
Then,  in Section 3 we give an overview of the narwhal data available for the analysis and the results of the application of our models to this data. The last section is dedicated to a discussion about these results, and potential extension of the models.

@@

The dynamic of the true positions of the narwhals is controlled by a stochastic differential equation. First, we recall the standard diffusion model in this context, and then we modify it to include the effect of the shore on the movement and constrain the position within a polygon.

@@

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
In this section, we obtain explicit distributional results about the process $X$. Though this model is considered in several papers, we found no mention of the closed form formulas derived here. In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, both the velocity and the position processes have a closed form formula. \\

@@

By integrating the velocity equation over time one gets the following conditional distribution for the location process :
\begin{equation}
	X(t+\Delta) \vert X(t)=x, V(t)=v \sim \mathcal{N}\left(x+\mu \Delta+A^{-1} \left( I_2-\exp(-A\Delta)\right)(v-\mu),Q(\Delta)\right)  
\end{equation}
Here, the exact formula for the covariance matrix $Q(\Delta)$ is the following :

\[
Q(\Delta)=\left( \Delta-2 \frac{\omega \sin(\omega \Delta)-\frac{1}{\tau} \cos(\omega \Delta)}{\frac{1}{\tau^2}+\omega^2 } \exp\left( -\frac{\Delta}{\tau} \right) +\frac{\tau}{2} \left( \frac{\omega^2-\frac{3}{\tau^2}}{\frac{1}{\tau^2}+\omega^2}-\exp\left( -\frac{2\Delta}{\tau}\right)\right) \right) I_2
\]
The location process is not markovian, but the whole process $(x(t), v(t))^\top$ is.
Both components of the location and velocity processes are independent.
and one can check that for $\omega=0$, the formulas that are derived here match the ones of the continuous-time correlated random walk model \cite{johnson_continuoustime_2008}. Figure~\ref{fig: sample RACVM} shows two samples of a RCVM along with the histograms of the empirical velocity norm. Derivations of the formulas given in this paragraph and the next one are detailed in appendix~\ref{section: detailed RACVM computations}.\\

@@

\subsubsection{State-space model}
\label{section: state space model}
In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}. We suppose here that $n$ positions are observed at discrete times $t_j$, $j \in \{1,\cdots, n\}$. $Y_j$ is the random vector representing the observed position at time $t_j$, while $X_j=X(t_j)$ is the true position at time $t_j$. The statistical model is the following:

@@

We consider parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time and make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that these parameters are constant on each time step $[t_j,t_{j+1}]$, equal to their value at time $t_j$. Denote $\Delta_j=t_{j+1}-t_j$, $V_j=V(t_j)$, $X_j=X(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix} 
	\frac{1}{\tau_j} & -\omega_j \\
	\omega_j & \frac{1}{\tau_j}
\end{pmatrix}$.

\begin{equation}
	\forall j \in \{1, \cdots, n\}, \quad 
	\left\{ \begin{array}{l}
		V_{j+1}=\mu_j +\exp(-A_j\Delta_j)(V_j-\mu_j)+\zeta_j \\
		X_{j+1}=X_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(_Vj-\mu_j)+\xi_j
	\end{array}\right. \\
	\label{eq: RCVM autoregressive formula}
\end{equation}

@@

Again, one can check that for $\omega_j=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.
This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter. This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} for the CVM, which corresponds to the special case $\omega=0$. Therefore, we extend this inference method to the more general model~\ref{eq: RACVM equation}.

@@

Here, $A(X(t),V(t))=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega(t) \\
	\omega(t) & \frac{1}{\tau}
\end{pmatrix}$
where $\omega(t)=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)=d(X(t),\partial\mathcal{D})$ is the distance to the boundary of the domain,
$\Theta(t)=\widehat{(X(t)-p(t),V(t))}$, $p(t)$ is the projection of $X(t)$ on $\partial \mathcal{D}$ and $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \Theta(s)\simeq \Theta(t)\]
In this case, the formulas of section~\ref{section: RACVM} are still valid.
The question still hanging is : what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation. It should always remain significantly lower than $D^{shore}$ for the process to be constrained. Moreover, a rotation needs to be considered only near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the narwhal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate,
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case \cite{johnson_continuoustime_2008}.
A weight based on the distance to the shore can be defined as follows :
\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

Based on this weight, a relationship between $f_{\omega}$ and the angle $\Theta$ and the distance to the shore $D^{shore}$ can be formulated as follows :

\begin{equation}
f_{\omega}(\Theta,D^{shore})=\frac{\omega_0}{2}\left(\tanh(\lambda(\theta+\theta^{*}))+\tanh(\lambda(\theta-\theta^{*})))\right)\times c(D^{shore})
\label{eq: smooth omega}
\end{equation}
for some $\omega_0 \in \R$, $\lambda>0$, $\theta^* \in [0,2\pi]$.

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. The general model is very flexible, and different species-specific behaviours could be described. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. \\
Figure~\ref{fig: trueomega} shows a smooth function obtained from equation~\ref{eq: smooth omega} for the parameter $\omega$, while figure~\ref{fig: splineomega} represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions. In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM). A sample of a CRCVM with smooth function $f_{\omega}$ defined as y equation~\ref{eq: smooth omega} is illustrated in Figure~\ref{fig: samples_fjords_CRCVM}. The domain in which the process $X$ is constrained represents Scoresby-Sound fjords system coastline in South-East Greenland.

\begin{figure}[H]
	\centering
		\centering
		\includegraphics[scale=0.25]{images/simulation study/true_smooth_omega_rect.png}
		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
		$D_0=0.3$ km, $\lambda=1.5$, $\kappa=0.5$ and $\omega_0=\pi$ rad/min}
		\label{fig: trueomega}
\end{figure}

\begin{figure}[H]
		\centering
		\includegraphics[scale=0.25]{images/simulation study/spline_smooth_omega_rect.png}
		\caption{$\omega$ (unit : $rad \cdot h^{-1}$) as a smooth function of $\Theta$ and $D_{shore}$. Values of $D_{shore}$ are in $km$ and values of $\Theta$ are in $rad$.
		This smooth function is constructed via bivariate tensor spline using a cubic spline basis with three degrees of freedom}
	\label{fig: splineomega}
\end{figure}

\begin{figure}[H]
		\centering
		\includegraphics[scale=0.7]{images/simulation study/illustrative_sample_fjords.png}
	\caption{Sample of a CRCVM. x and y axis are in metres and initial positions are indicated by a red cross and red points represent the nearest points on the boundary of the domain. One  five day sample with time step $h=5$ min within the spatial constraints of Scoresby Sound fjords system in South-East Greenland. Parameters are $\nu=4$ km/h and $\tau=1$ h. }
	\label{fig: samples_fjords_CRCVM}
\end{figure}

\subsection{Simulation study}

@@

\subsubsection{Simulation of constrained motion}
\label{section: simulation constrained motion}
In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$.\\

100 batches of 6 individual trajectories over $12$ hours with time step $h=1$ min were sampled in a rectangular domain with sides of length $8$ km from a CRCVM. Gaussian noise with standard deviation $\sigma_{obs}=10$ m is considered to represent measurement error and approach a practical case. Denoting $n$ the number of observations for each individual trajectory, $Y_{ij}$ the observation with noise and $X(t_{ij})=X_{ij}$ the true location, the precise statistical set up is the following:

\begin{equation}  \forall  i \in \{1,\cdots,6\}, \quad  \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots, n\} \\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) \\
		dX_i(t)=V_i(t) dt \quad \\
		dV_i(t)=-A(X_i(t),V_i(t))V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t)  \\
		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}, \\\
		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}, \\
		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
		\omega_i(t)=f_{\omega}(D^{shore}_i(t),\Theta_i(t)), \quad t \geq 0\\
		f_{\omega}(D^{shore},\Theta)=\omega_0+\sum_{k=1}^{L} \omega_k \psi_k(D_{shore},\Theta), \quad (D_{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: rect simulation parameters}
\end{equation}
Here, $L=8$ and $\psi_k$ are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
1.8$ construction of tensor splines basis function. We choose $\sigma_{\tau}=0.1$ h and $\sigma_{\nu}=0.3$ km/h to represent the individual variability, wheras $\tau_0$ and $\nu_0$ are fixed respectively to $1$ h and $4$ km/h.
For each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary.   Of the $600$ simulated trajectories, none reach the boundary.\\

100 batches of 6 individual trajectories over $24$ hours were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints in our application. The parameters were set as in~\ref{eq: rect simulation parameters}, with only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ changed to adapt to this specific domain.
Again, for each trajectory, initial velocity was set to $0$ and initial position was uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore. Time steps were also kept constant, equal to $1$ min. None of the $600$ trajectories reached land. Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often. For the fjords system, this already happens if we try to sample directly with $5$ min time steps. Constrained samples within the geometry of Scoresby sound fjords are shown in~\ref{fig: samples_fjords_CRCVM}. Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.

\subsubsection{Parameters estimation with \texttt{smoothSDE}}

Estimation with $\texttt{smoothSDE}$ relies on the state-space formulation in section~\ref{section: state space model} with the approximation that the parameters are constant on each time step between consecutive observations.
The complete likelihood is computed from the observations $Y_{ij}$ as a by-product of the Kalman filter algorithm \cite{michelot_varying-coefficient_2021}.
Laplace's approximation of the integral of this complete likelihood over the random effects is computed using $\texttt{TMB}$ \cite{kristensen_tmb_2016} and optimization is performed via the \texttt{optim} function in $\texttt{R}$ with the BFGS gradient method. We refer to \cite{wood_generalized_2017} section $6.6$ for the specification of the regularization terms based on a wiggliness measure in the mixed effect models involving gaussian priors for the spline coefficients $\omega_k$, $k \in \{1,\cdots,L\}$. \\

$100$ batches of noisy samples obtained as described in section~\ref{section: simulation constrained motion} are used as data for estimation, both in the rectangular domain and in Scoresby Sound fjords system. We thus obtain $100$ parameters estimate in the two cases. The covariates $D^{shore}$ and $\Theta$ are computed from the observed positions and the smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}. The marginal spline degree of freedom was set to $3$ for each covariate, which corresponds to one single knot in each dimension. Hence, $9$ coefficients (including the intercept) are to be estimated for the smooth function $\omega$, plus the two parameters $\tau$ and $\nu$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$. Measurement error $\sigma_{obs}$ is supposed to be known. 
We fixed initial parameter values to $\tau_0=1$ h, $\nu_0 = 1$ km/h,$\omega_k=0$ for all $k \in \{0,\cdots,L\}$. Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories for different values of the distance to the shore in the rectangular and the fjords domain.

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q4_ExpShore .png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_rect_hf_ne1_omega_theta_q2_ExpShore .png}
		\caption{}
	\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q4_ExpShore .png}
	\caption{}
\end{subfigure}
\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/simulation study/fe_crcvm_fjords_hf_ne1_omega_theta_q2_ExpShore .png}
	\caption{}
\end{subfigure}
	\caption{Estimates of the smooth parameter $\omega$ as a function of $\Theta$ only,. The red curve represents the true value of $\omega$. (a)-(b) Estimation from samples in the rectangular domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances. (c)-(d) Estimation from samples in the fjords domain with distance to shore equal to the $25\%$ and $75\%$ quantiles of observed distances. }
\label{fig: marginal_estimates_CRCVM}
\end{figure}
